Simple repression draft

### 2020-05-27 doc/section_04_bayesian_inference.tex
MRM rewrote Bayesian inference (unreg promoter) with more transparent language

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}, a one-state bursty model of
constitutive promoters, achieves a Goldilocks level of complexity. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on
single-cell mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this is in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig2:constit_cartoons}.  These models have Fano factors $\nu\le 1$
meaning they are underdispersed relative to the Poisson distribution. We will also not explicitly
consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

In performing parameter inference on the single-molecule data
of~\cite{Jones2014}, building a generative model is straightforward. The data
are single-cell mRNA counts, and we assume different cells to be independent and
identically distributed from the underlying generative model. The generative
model for our case is the steady-state mRNA distribution computed from the
master equation under consideration.

\subsubsection{Model 1: Poisson promoter}
% \mmnote{A lot of the details I've written here might be excessive for main text and serve better as an intro-by-example to Bayes in the appendix, then reduce main text to a bit more bare bones?}
For this model the master equation of interest is
Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
\deriv{t}p_{m,U}(t) = rp_{m-1,U}(t) - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) - \gamma p_{m,U}(t),
\end{equation}
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$. This steady state distribution defines the
likelihood $p(m\mid\lambda)$ for our Bayesian inference, i.e. 
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}
\label{eq:poisson_inference010}
\end{equation}
is the probability of observing a single cell with $m$ mRNAs given a value of
$\lambda$, written to make the conditioning explicit.

With our likelihood in hand, we can invoke Bayes theorem to write down the
posterior distribution on our model parameter $\lambda$, given by
\begin{equation}
p(\lambda\mid D) \propto p(D\mid\lambda) p(\lambda),
\end{equation}
where our data $D=\{m_1, m_2,\dots, m_N\}$ is a list of mRNA counts over a
population of $N$ identical cells. As is standard we have neglected the factor
$p(D)$ on the right hand side since it is independent of $\lambda$ and serves
only as a normalization factor. Since we assume each cell's mRNA count is
independent of others, the likelihood of the full inference problem
$p(D\mid\lambda)$ is simply a product of the single cell likelihoods given by
\eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}
This is often notated simply as
\begin{equation}
D \sim \text{Poisson}(\lambda),
\end{equation}
which is read as the data $D$ is Poisson distributed with mean $\lambda$. The
$\sim$ operator literally denotes ``is distributed according to.'' The
conditioning on $\lambda$ is no longer explicitly notated, only implied by its
appearance on the right hand side.

To proceed we need to specify a prior. In this case we are extremely data-rich,
as the dataset from Jones et.\ al~\cite{Jones2014} has of order 1000-3000
single-cell measurements for each promoter, so our choice of prior matters
little here, as long as it is sufficiently broad. For details on the prior 
selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
purpose here it suffices to specify that we use a prior of the form $\lambda
\sim \text{Gamma}(\alpha, \beta)$. Putting a gamma prior on $\lambda$ introduces
two new parameters $\alpha$ and $\beta$ which parametrize the gamma distribution
itself, which we use to encode the range of $\lambda$ values we view as
reasonable. Recall $\lambda$ is the mean steady-state mRNA count per cell, which
\textit{a priori} could plausibly be anywhere from 0 to a few hundred.
$\alpha=1$ and $\beta=1/50$ achieve this, since the gamma distribution is
strictly positive with mean $\alpha/\beta$ and standard deviation
$\sqrt{\alpha}/\beta$.

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior
known as the \textit{conjugate} prior for a Poisson likelihood allows us to 
get a closed form for the posterior distribution $p(\lambda \mid D)$. In 
particular it can be shown that~\cite{Gelman2013}
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right),
\end{equation}
where we defined the sample mean $\bar{m} = \frac{1}{N}\sum_k m_k$ for
notational convenience. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to make the 
excellent approximation 
\begin{equation}
\lambda
\sim \text{Gamma}\left(\alpha + \bar{m}N, \beta + N\right)
\approx \text{Normal}\left(\bar{m}, \sqrt{\frac{\bar{m}}{N}}\right).
\end{equation}
As an example with real numbers, for the \textit{lacUV5} promoter, Jones et.\
al~\cite{Jones2014} measured 2648 cells with an average mRNA count per cell of
$\bar{m} \approx 18.7$. In this case then, our posterior is
\begin{equation}
\lambda
\sim \text{Normal}\left(18.7, 0.08\right),
\label{eq:posterior_poisson}
\end{equation}
which suggests we have inferred our model's one parameter to a precision of
order 1\%.

We alert the reader not to confuse a tightly constrained posterior distribution
with a good generative model. In other words, not because the single parameter
for our Poisson likelihood can only take a small range of values constrained by
the data, this implies that such Poisson distribution characterizes the
underlying cell-to-cell variability we observe in the data. To systematically
check the validity of the generative model we can make use of \textit{posterior
predictive checks}. This consists in generating synthetic data using our
generative model for many plausible values of our parameter. For the case of our
\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
samples drawn from a Poisson distribution with  $\lambda$ chosen appropriately. We then repeat
this procedure 1000 times. Clearly if those data appear quite different, the
model has a problem. Put another way, if we suppose the generative model is
true, then the synthetic datasets we generate should resemble the actual
observed data, and if not, it suggests the model is missing important features.

To compare so many samples with the actual observed data, we can use empirical
cumulative distribution functions (ECDF) of the distribution quantiles. This 
representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
example, the median for each possible mRNA count for the Poisson distribution
is shown as a dark green line, while the lighter green contains 95\% of the
posterior predictive samples. This way of representing the posterior predictive
checks gives us a sense of the range of data we might consider plausible, under
the assumption that the model is true. In this case it is quite obvious that the
observed data, plotted in orange, could not plausibly come from this Poisson
generative model. We will cover the other model shown in
Figure~\ref{fig:constit_post_full}(B) next.

\subsubsection{Model 5 - Bursty promoter}
Let us now consider the problem of parameter inference from FISH data for model
five from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate nondimensionalized
by the mRNA degradation rate $\gamma$. As sketched earlier, the story of this
distribution is of geometrically-distributed bursts of mRNA, where the arrival
of bursts is a Poisson process with rate $k_i$ and the mean size of a burst is
$b$.

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood --the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one to two.
Details on the arguments behind our prior distribution selection are left for
Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2) \\
\ln b \sim \text{Normal}(0.5, 1) \\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
Note that we specify a distribution over the logarithm of our parameters $k_i$
and $b$. This is commonly the case for parameters for which our previous 
knowledge based on our domain expertise spans several orders of magnitude.

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in~\fig{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

Flush with success having found the model's posterior to be well-identified, the
next step is posterior predictive sampling. The only procedural difference
compared to the Poisson promoter model is that, since we have no analytical
posterior expression but only samples from the posterior, we generate a
synthetic dataset (i.e., posterior predictive samples) for each posterior sample
from the MCMC sampling. \fig{fig:constit_post_full}(B) shows the resulting
predictive ECDF; similarly to the Poisson promoter model, the median posterior
predictive ECDF is plotted as a dark blue line, while a lighter blue shaded
region encompasses 95\% of the posterior predictive samples. Unlike the Poisson
promoter model, the experimental ECDF closely tracks the posterior predictive
ECDF, indicating this model is actually able to generate the observed data and
increasing our confidence that this model is at least not wrong.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is
labeled with a unique number in both (C) and (D) for cross comparison.}
\label{fig:constit_post_full}
\end{figure}

\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
distribution from individually MCMC sampling all 18 constitutive promoter
datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
each posterior we find and plot the curve that surrounds the 95\% highest
probability density region. What this means is that each contour 
encloses approximately 95\% of the samples, and thus 95\% of the probability
mass, of its posterior distribution. Posterior predictive ECDFs (not shown)
display a similar level of agreement between data and predictive samples as for
the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
\mmnote{Might be worth showing some/all of these PPCs in supplement for
completeness? We could do postage stamp sized diff plots and probably fit all 18
promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}

@@

It is interesting to connect this observation to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state equilibrium model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m \right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r \ll 1)$) results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

>>>>

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves a Goldilocks level of complexity. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on
single-cell mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data in Figure~\ref{fig2:constit_cartoons}(B) has Fano factor
$\nu>1$, we might already suspect that this model is incapable of fitting the
data. We will verify that this is in fact the case. Using the same argument we
can immediately rule out models 2 and 3 from
Figure~\ref{fig2:constit_cartoons}(A). These models have Fano factors $\nu\le 1$
meaning they are underdispersed relative to the Poisson distribution. We will
also not explicitly consider model 4 from~\fig{fig2:constit_cartoons}(A) since
it was already thoroughly analyzed in~\cite{Razo-Mejia2020}, and since model 5
can be viewed as a special case of it.

Our objective for this section will then be to assess whether or not model 5 is
quantitatively able to reproduce experimental data. In other words, if our claim
is that the level of coarse graining in this model is capable of capturing the
relevant features of the data, then we should be able to find values for the
model parameters that can match theoretical predictions with single-molecule
mRNA count distributions. A natural language for this parameter inference
problem is that of Bayesian probability. We will then build a Bayesian inference
pipeline to fit the model parameters to data. To gain intuition on how this
analysis is done we will begin with the ``wrong'' model 1 in
Figure~\ref{fig2:constit_cartoons}(A). We will use the full dataset of
single-cell mRNA counts from~\cite{Jones2014} used in
Figure~\ref{fig2:constit_cartoons}(B).

\subsubsection{Model 1: Poisson promoter}

For this model the master equation of interest is
Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
{d\over dt}p_{m,U}(t) = 
        rp_{m-1,U}(t) 
        - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) 
        - \gamma p_{m,U}(t),
\end{equation}
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$~\cite{Sanchez2013}. The goal of our inference 
problem is then to find the probability distribution for the parameter value
$\lambda$ given the experimental data. By Bayes theorem this can be written as
\begin{equation}
p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
\end{equation}
where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
counts. As is standard we will neglected the denominator $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as a
normalization factor.

The steady-state solution for the master equation defines the likelihood
term for a single cell $p(m \mid \lambda)$. What this means is that for a given
choice of parameter $\lambda$, under model 1 of
Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
single cell with probability
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
\label{eq:poisson_inference010}
\end{equation}
Assuming each cell's mRNA count in our dataset is independent of others, the
likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
of the single cell likelihoods given by \eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}

To proceed we need to specify a prior distribution $p(\lambda)$. In this case we
are extremely data-rich, as the dataset from Jones et.\ al~\cite{Jones2014} has
of order 1000-3000 single-cell measurements for each promoter, so our choice of
prior matters little here, as long as it is sufficiently broad. For details on
the prior selection we refer the reader to Appendix~\ref{sec:bayesian}. For our
purpose here it suffices to specify that we use as prior a Gamma distribution.
This particular choice of prior introduces two new parameters, $\alpha$ and
$\beta$, which parametrize the gamma distribution itself, which we use to encode
the range of $\lambda$ values we view as reasonable. Recall $\lambda$ is the
mean steady-state mRNA count per cell, which \textit{a priori} could plausibly
be anywhere from 0 to a few hundred. $\alpha=1$ and $\beta=1/50$ achieve this,
since the gamma distribution is strictly positive with mean $\alpha/\beta$ and
standard deviation $\sqrt{\alpha}/\beta$.

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior known
as the \textit{conjugate} prior for a Poisson likelihood. This type of priors
have nice properties to work with, in particular a conjugate prior allows us to
get a closed form for the posterior distribution $p(\lambda \mid D)$. In
particular it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. In particular the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
cells n our dataset. Furthermore given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we got such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, the parameter 
$\lambda$ for our model is highly constrained to take a narrow range of values,
but a narrow posterior distribution does not speak to the goodness of fit of the
model. As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed there is not good correspondence
when contrasting the Poisson model with the experimental .

\subsubsection{Model 5 - Bursty promoter}

Let us now consider the problem of parameter inference for model five
from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
binomial distribution in terms of an intuitive story, we imagine the arrival of
bursts as a Poisson process with rate $k_i$, where each burst has a 
geometrically-distributed size with mean size $b$.

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use when stating Bayes
theorem. Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
m_N\}$, here Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood -- the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom} -- so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one parameter to two. Details on the arguments behind our prior
distribution selection are left for Appendix~\ref{sec:bayesian}. We state here
that the natural scale to explore these parameters is logarithmic. This is
commonly the case for parameters for which our previous knowledge based on our
domain expertise spans several orders of magnitude. For this we chose log-Normal
distributions for both $k_i$ and $b$. Details on the mean and variance of these
distributions can be found in Appendix~\ref{sec:bayesian}.

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in Figure~\ref{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

Having found the model's posterior to be well-identified as with the Poisson
promoter, the next step is to compare both models with experimental data. To do
this for the case of the bursty promoter, for each of the parameter samples
shown in Figure~\ref{fig:constit_post_full}(A) we generated negative
bionomial-distributed mRNA counts. As MCMC samples parameter space
proportionally to the posterior distribution, this set of random samples span
the range of possible values that we would expect given the correspondence
between our theoretical model and the experimental data. A similar procedure can
be applied to the Poisson promoter. To compare so many samples with the actual
observed data, we can use empirical cumulative distribution functions (ECDF) of
the distribution quantiles. This representation is shown in
Figure~\ref{fig:constit_post_full}(B). In this example, the median for each
possible mRNA count for the Poisson distribution is shown as a dark green line,
while the lighter green contains 95\% of the randomly generated samples. This
way of representing the fit of the model to the data gives us a sense of the
range of data we might consider plausible, under the assumption that the model
is true. For this case, as we expected given our premise of the Poisson promoter
being wrong, it is quite obvious that the observed data, plotted in black. An 
equivalent plot for the bursty promoter model is shown in blue. Again the darker
tone shows the median, while the lighter color encompasses 95\% of the randomly
generated samples. Unlike the Poisson promoter model, the experimental ECDF
closely tracks the posterior predictive ECDF, indicating this model is actually
able to generate the observed data and increasing our confidence that this model
is at least not wrong.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is labeled with a unique number in
both (C) and (D) for cross comparison and for comparison with 
Figure~\ref{fig2:constit_cartoons}(B).}
\label{fig:constit_post_full}
\end{figure}

\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
distribution from individually MCMC sampling all 18 constitutive promoter
datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
each posterior we find and plot the curve that surrounds the 95\% highest
probability density region. What this means is that each contour 
encloses approximately 95\% of the samples, and thus 95\% of the probability
mass, of its posterior distribution. Theory-experiment comparisons (not shown)
display a similar level of agreement between data and predictive samples as for
the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
\mmnote{Might be worth showing some/all of these PPCs in supplement for
completeness? We could do postage stamp sized diff plots and probably fit all 18
promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}

@@

It is interesting to connect this observation to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state equilibrium model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m
\right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.
